\section{Implementations of matchertext}
\label{sec:impl}

This section is mainly a placeholder at the present.
Some preliminary work is underway
implementing experimental matchertext extensions
for several languages and embedding-oriented syntaxes.
This section is intended to be expanded
as we gain experience
implementing and using matchertext extensions.

For those wishing to help with implementation and experimentation,
the following are some of the key work items
enabling us to start experimenting with matchertext in the context
of any particular language of interest:
\begin{itemize}
\item	A brief specification of proposed
	syntactic extensions for hosting matchertext
	and/or conveniently writing embedded matchertext,
	adapted to the specific language of interest,
	with clearly-defined rationale for the particular syntax choices.
\item	An implementation of those extensions for hosting and/or embedding,
	selectively enabled via configuration parameters
	for backward compatibility,
	in some mature processor for the language of interest
	(\eg a compiler, interpreter, or library implementation).
\item	A configurable extension to the language processor
	that optionally causes it
	to check and enforce the matchertext discipline
	in processed text:
	\ie to verify that matchers match in source files or strings.
\item	Purely for experimentation purposes,
	an extension of some processor for the language
	that can analyze ``legacy'' source files in the language
	to detect and categorize matchertext violations
	(\eg whether in string literals, comments, or elsewhere).
	We will use this to help estimate the likely ``pain''
	of adopting the matchertext discipline in the language
	and how commonly this pain would affect typical code today.
\end{itemize}
